### 2018-07-23 express2018/intro.tex
Fixed 'p. 2, l.1: I don’t know the “Hennessy and Deng” lemmas; put a reference.' #30

The main goal of the work described in this paper is a rather
comprehensive formalisation  of the core of the theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions of contractions.
The formalisation however is not confined to the theory of  unique solutions, but embraces 
most of the 
core of the theory of CCS \cite{Mil89}
(partly because the theory of unique solutions
 relies on a number of more fundamental results):
indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), 
their basic algebraic theory, various versions of ``bisimulation up to''
techniques (e.g., bisimulation up-to bisimilarity),
the main properties  of the rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and of the expansion preorder (an
efficiency-like refinement of bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy and Deng lemmas, and two proofs that rooted bisimilarity is the largest
congruence included in bisimilarity:
one as in Milner's
book,  requiring the hypothesis that  no processes can use all labels; the other without
such hypothesis, essentially formalising van Glabbeek's paper \cite{vanGlabbeek:2005ur}
(such a proof however follows the structure of the ordinal numbers, which cannot be handled
in HOL, and therefore it is restricted to finite-state processes ).
Similar theorems are proved for the rooted contraction preorder.
In this respect,  the work is an extensive experiment with the use of the HOL theorem prover and its
most recent developments, including a package  for expressing coinductive definitions.
The
work consists of about 20,000 lines of proof scripts in Standard ML.

From the CCS theory viewpoint, the formalisation has offered us the possibility of 
further refining the theory of unique solutions. 
In particular, the existing theory  placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.  
We have thus refined the proof  technique based on contractions by moving to  
\emph{rooted contraction}, that is, the coarsest (pre)congruence contained in the contraction
preorder.  The resulting 
 unique-solution theorem is valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

>>>>

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core of the theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions of contractions.
The formalisation, however, is not confined to the theory of unique
solutions of equations, but embraces a significant portion the theory of CCS \cite{Mil89}
(mostly because the theory of unique solutions relies on a large number of more fundamental results).
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g. the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g., bisimulation up-to bisimilarity), and of the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying the rooted bisimilarity is the largest (coarsest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{vanGlabbeek:2005ur}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover and its
most recent developments, including a package for expressing coinductive definitions.
In total, the work consists of about 20,000 lines of proof scripts in Standard ML.

From the viewpoint of CCS theory, the formalisation has offered us the possibility of 
further refining the theory of unique solutions of equations.
In particular, the existing theory placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
We have thus refined the contraction-based proof technique, by moving to  
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.
